Monte Carlo tree search compared to A\(^*\) in airspace configuration decision problem

### 2018-05-18 article/article.tex
(+) Added other figure
(~) Small corrections

\begin{abstract}
  Dynamic airspace configuration falls under highly combinatorial partitioning
  problems. Since exact methods tend to be overwhelmed by the complexity of such
  problems, a stochastic approach is here considered. This paper presents the
  the Monte Carlo tree search algorithm using UCT search and adapted to one
  player games. The application of the algorithm to the airspace partitioning is
  then detailed. Finally the Monte Carlo tree search is compared to the A\(^*\).
\end{abstract}

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(C_{tr}(P_1, P_2)\): cost of transition between partitions \(P_1\) and
    \(P_2\);
  \item \(\pi = [P_1, \dots, P_n]\): a sequence of partitions (called \emph{path}
      in the tree search algorithms);
  \item \(f(\pi)\): cost associated to sequence of partitions \(\pi\);
  \item \(\mathcal{N}\): the set of nodes;
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item $h \colon \mathcal{N} \to \mathbb{R}^+$: heuristic estimating the cost
    from a node \(u\) to a final node;
  \item \(\mu_v\): mean value of outcomes of simulations run through or from a
    node \(v\);
  \item \(T_v\): number of simulations run through or from a node \(v\);
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

Currently, configuration is mainly decided on the fly by chief officer.
This decision is based on the actual workload on each position. In this
approach, future workload estimation is based on the controller's feelings and
it therefore lacks a workload prediction tool.
This papers aims at providing this tool as well as a method offering a smoother
workload during the day.

@@

In this paper, a temporal sequence of configuration is provided, as done
in~\cite{treimuth2016branch} or~\cite{sergeeva2017dynamic}.
Two costs are considered to create the sequence, namely the cost
associated to each configuration and a transition cost. The former is based on
the workload estimated for one sector given a neural network
(see~\cite{gianazza2010forecasting}).

The sequence of configurations is represented as a tree. The resulting
sequence will therefore be a path from the root to a leaf of this tree. This
problem is highly combinatorial (partitioning of the airspace). Since stochastic
tree search algorithms, combined with deep neural networks have proved
themselves worthy by outranking the best human player of one of the most highly
combinatorial game which is the game of Go, the Monte Carlo tree
search~\cite{browne2012survey} algorithm is used in this paper to fulfill the
previously mentioned task.

@@

The dynamic airspace configuration problem requires a model of the airspace. As
done in~\cite{sergeeva2017dynamic} or~\cite{treimuth2016branch} the airspace
is modelled via a graph. In those graphs, vertices represent elementary modules
and an edge links two adjacent modules. In~\cite{treimuth2016branch}, to be
able to produce a sequence of configurations, the graph is time dependent.
An other way to model the problem is to use a constrained set of
configurations as in~\cite{gianazza2010forecasting} and~\cite{bloem2010dynamic}.
This way any configuration will match specified requirements, which can be
qualified hard constraints.

In most cases the cost of a configuration is based on the workload. Each
approach seems to give their own representation of the workload. For instance,
\cite{bedouet2016towards} determined workload density proportionally to the time
spent by aircraft in each sector. A simpler version~\cite{sergeeva2017dynamic}
only uses the number of aircraft. On the other hand, more complex methods,
involving many more inputs are also available. For
instance,~\cite{gianazza2010forecasting} used several indicators, such as sector
volume, or vertical incoming flows in the next 15 and 60 minutes. Those
indicators ease the training of a neural network forecasting the workload.

Other soft constraints appear to be relevant to have a better model of
the problem. For instance in~\cite{sergeeva2017dynamic}
and~\cite{bedouet2016towards} a
coordination cost is defined. It represents the surplus of work added by
flights travelling from one sector to an other. The shape of the
resulting sector is considered, as the simpler is
the shape, the easier it is to manage. Complex shapes are therefore avoided,
thanks to the notion of compactness in~\cite{jagare2013airspace} and balconies
in~\cite{sergeeva2017dynamic}. To smooth the transition between two
configurations, the work associated with the reallocation of one or more modules
is evaluated. This can be used in a cost function aimed to be minimised
(in~\cite{bedouet2016towards}).

@@

The Monte Carlo tree search algorithm has been quickly used in two players game.
Only three years after its apparition in 1990, Br\"ugmann applies it to the Go
game in~\cite{brugmann1993mcgo}. While the Monte-Carlo tree search is still
extensively used in two players games (and especially the Go game), its
adaptation to one player game has been worked on. Auer \textit{et al.} propose
an
upper confidence bound formula which appears to be more efficient for one player
games as seen in~\cite{sebag2010fuse}. The introduced formula uses the standard
deviation of the outcome of the simulations.
The latter article also uses the rapid
action value estimation (RAVE) technique to quicken the convergence of the
algorithm. RAVE uses the ``all moves as first'' heuristic, in which all
moves\footnote{a move is informally considered as a decision taken regarding
which state to choose while descending the search tree}
seen during simulations are considered as a first move. This allows the
algorithm to update more statistics in one simulation. Gelly \textit{et al.} use
in~\cite{gelly2012grand} the RAVE technique coupled with several heuristics. The
heuristics bias the initialisaton of a node in the search tree, pre filling its
statistics using prior knowledge of the problem.

@@

The task of building an optimal sequence of partitions (lowering
as much as possible the workload of each controller) is fulfilled by a
stochastic tree search method, namely the Monte Carlo tree search. To assess the
exactness of the results, an exact method (here A\(^*\)) is used.

@@

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then the MCTS algorithm is called back with
node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{\mu_v | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{N}^2 \to \mathbb{R}\). The
function \(\first(G)\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \first(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

@@

In this section, we discuss the model established to approach this configuration
problem. First we defined a structure for a controled sector and all possible
transitions through time. Then a workload model is defined. And finally we
determined a cost per partition and a cost function aimed to be minimized.

@@

In the configuration problem, the airspace needs to be divided into sectors. Each
sector represents an area controlled by two controllers and is composed of
one or more elementary modules. A partition $P$ is a set of
sectors covering all the airspace.

A sector is a group of elementary modules, but in practice only a collection
of sectors are allowed in this model. This collection is referred as the context.
This restriction reflects the fact the more complex the shape is, the harder it
is to manage.

@@

During the day, and with the evolution of the trafic, the workload (rigorously
defined in~\ref{sub:workload}) vary. To help controllers to maintain a
homogeneous level of workload, the airspace partitioning needs to be adapted
through a transition from an inital partition to a target partition.
This maneuver requires either coordination or the opening of a new position
and hence increases controller's workload.
This is the reason why a full reconfiguration isn't feasible, and
only a subset of all available transitions can be operated.

In the model considered here, three transistions are possible namely a merge,
a transfer
or a split. For an airspace composed of three modules $A$, $B$, and $C$,
those actions represent:
\begin{itemize}
  \item merge: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A, B, C\}\}$
  \item split: $\{\{A, B, C\}\} \quad \rightarrow \quad \{\{A, B\}, \{C\}\}$
  \item transfer: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A\}, \{B, C\}\}$
\end{itemize}

Each transition is composed of at most one merge, split or tranfer; more would
result in too complex reconfigurations.

@@

Let us now precise the model established to evaluate a workload metric. The
workload depends on external criteria such as sector and trafic complexity
or humain factors (e.g. controller health or stress). In the model assumed here,
the workload felt by a controller is directly and only related to the trafic
complexity.

The simplest approch considers only the number of aircraft per sector $N_\text{aircraft}$. We
defined two arbitary thresholds, namely $th_\text{low}$ and $th_\text{high}$, defining
three zones where the workload is low (resp. normal and high) if $N_\text{aircraft} < th_\text{low}$
(resp. $th_\text{low} < N_\text{aircraft} < th_\text{high}$ and $th_\text{high} < N_\text{aircraft}$).
Those workload levels are translated to probabilities $p_\text{low}$, $p_\text{normal}$ and
$p_\text{high}$, where $p_\text{low}(S)$ is the probability that the sector $S$ is underloaded
(resp. balanced and overloaded). In short, for a sector $S$ at time $t$ and $N_\text{aircraft}$:
\begin{equation}
  \begin{aligned}
    p_\text{low}^{S, t} &= \mathbb{1}_{[0, th_\text{low}]}(N_\text{aircraft}) \\
    p_\text{normal}^{S, t} &= \mathbb{1}_{[th_\text{low}, th_\text{high}]}(N_\text{aircraft})\\
    p_\text{high}^{S, t} &= \mathbb{1}_{[th_\text{low}, \infty[}(N_\text{aircraft})
  \end{aligned}
\end{equation}
where $\mathbb{1}$ is the indicator function defined by:
\begin{equation}
  \mathbb{1}_{I}(x) =
  \begin{cases}
    1 & \text{if\,} \quad x \in I\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

This expression of the workload, expressed in term of probabilities, allow us to
use a more complex model for the workload such as a neural network
(see~\cite{gianazza2010forecasting}). This method learns the probabilities $p_\text{low}$,
$p_\text{normal}$ and $p_\text{high}$ based on parameters including trafic complexity and
the complexity of the sector (e.g. airspace volume).

@@

In order to evaluate a given partition $P$ at time $t$, a cost $C(P, t)$ needs
to be defined. In this paper, this cost depends on the workload in each the
sector. The definitions given in~\cite{ferrari2017} are used to represent a high
(respectively normal and low) cost for partition \(P\):
\begin{equation}
  \begin{aligned}
    c_+(P, t) &= \sum_{S \in P_t} \delta_{h}(S, t) \cdot p_\text{high}^{S, t} \cdot |S|^2\\
    c_=(P, t) &= \left(
    \sum_{S \in P_t} \delta_{n}(S, t) \cdot p_{norm}^{S} \cdot |S|^{-2}
    \right)^{-1}\\
    c_-(P, t) &= \sum_{S \in P_t} \delta_{l}(S, t) \cdot p_\text{low}^{S, t} \cdot |S|^{-2}
  \end{aligned}
\end{equation}
with $\delta_{h}(P, t)$ (resp. $\delta_{n}(P_t)$ and $\delta_{l}(P_t)$) equals 1
if the probability $p_\text{high}$ (resp. $p_\text{normal}$ and $p_\text{low}$) is superior to
the two others, and 0 otherwise.

@@

The parameters $\alpha$, $\beta$, $\gamma$ and $\lambda$ (all positive)
determine a priority on which parameter to optimize. For instance, a high
value $\alpha$ help to
minimize $c_+$, hence the overall number of sectors with too much trafic.
In a real application, it maybe interesting to order those parameters as
follow: $\alpha > \gamma > \beta > \lambda$.

@@

with $\theta > 0$ a parameter to determine.
This is the immediate function to minimize in the A$^*$ algorithm.

@@

The traffic is simulated the following way, each elementary module has an
associted list which, at index \(i\) has the number of aircraft in it at time
step \(i\). Thoses lists has been randomly generated. The set of lists form a
scenario which is used in all following tests.

@@

Considering paragraph~\ref{ssec:theoretical_performances}, the MCTS should
eventually converge to the optimal value. This result reveals a flaw in the
implementation.
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/longterm.data};
      \addplot+[dashed, domain=0:10] {1268};
    \end{axis}
  \end{tikzpicture}
  \caption{The continuous line shows the total cost of a path computed by the
    MCTS algorithm. The x-axis shows the time allowed to the algorithm. The
    dashed curve is 15\% of the optimal cost found with the \(A^*\)
  algorithm.}\label{fig:longterm}
\end{figure}

>>>>

\begin{abstract}
  Dynamic airspace configuration is a highly combinatorial partitioning
  problem. Since exact methods tend to be overwhelmed by the complexity of such
  problems, a stochastic approach is here considered. This paper presents the
  Monte Carlo tree search algorithm using UCT search and adapted to one
  player games. The application of the algorithm to the airspace partitioning
  problem is
  then detailed. Finally the Monte Carlo tree search is compared to the A\(^*\).
\end{abstract}

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: air traffic control sector, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(C_{tr}(P_1, P_2)\): cost of transition between partitions \(P_1\) and
    \(P_2\);
  \item \(\pi = [P_1, \dots, P_n]\): a sequence of partitions (called \emph{path}
      in the tree search algorithms);
  \item \(f(\pi)\): cost associated to sequence of partitions \(\pi\);
  \item \(\mathcal{N}\): the set of nodes;
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item $h \colon \mathcal{N} \to \mathbb{R}^+$: heuristic estimating the cost
    from a node \(u\) to a final node;
  \item \(\mu_v\): mean value of outcomes of simulations run through or from a
    node \(v\);
  \item \(T_v\): number of simulations run through or from a node \(v\);
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

Currently, configuration is mainly decided on the fly by the chief officer.
This decision is based on the actual workload on each position. In this
approach, future workload estimation is based on the controller's feelings and
it therefore lacks a workload prediction tool.
This papers aims at providing a method predicting the airspace configuration.

@@

In this paper, a temporal sequence of configurations is considered, as
in~\cite{treimuth2016branch} or~\cite{sergeeva2017dynamic}.
Two costs are considered to create the sequence, namely the cost
associated to each configuration and a transition cost. The former is based on
the workload estimated for one sector given by a simple model. More complex
models might be used such as a neural network
(see~\cite{gianazza2010forecasting}).

The sequence of configurations is extracted from a tree. The resulting
sequence will therefore be a path from the root to a leaf of this tree. This
problem is highly combinatorial (partitioning of the airspace). Since stochastic
tree search algorithms, combined with deep neural networks have proved
themselves worthy by outranking the best human player of one of the most highly
combinatorial game which is the game of Go, the Monte Carlo tree
search~\cite{browne2012survey} algorithm is used in this paper to fulfill the
previously mentioned task.

@@

The dynamic airspace configuration problem requires a model of the airspace.
In~\cite{sergeeva2017dynamic} or~\cite{treimuth2016branch} the airspace
is modelled via a graph. In those graphs, vertices represent elementary modules
and an edge links two adjacent modules. In~\cite{treimuth2016branch}, to be
able to produce a sequence of configurations, the graph is time dependent.
An other way to model the problem is to use a constrained set of
configurations as in~\cite{gianazza2010forecasting} and~\cite{bloem2010dynamic}.
This way any configuration will match specified requirements, which can be
qualified as hard constraints.

In most cases the cost of a configuration is based on the workload. Each
approach seems to give their own representation of the workload. For instance,
\cite{bedouet2016towards} determined workload density proportionally to the time
spent by aircraft in each sector. A simpler version~\cite{sergeeva2017dynamic}
only uses the number of aircraft. On the other hand, more complex methods,
involving many more inputs are also available. For
instance,~\cite{gianazza2010forecasting} used several indicators, such as sector
volume, or vertical incoming flows in the next 15 and 60 minutes. Those
indicators as inputs to a neural network forecasting the workload.

Other soft constraints appear to be relevant to have a better model of
the problem. For instance in~\cite{sergeeva2017dynamic}
and~\cite{bedouet2016towards} a
coordination cost is defined. It represents the surplus of work added by
flights travelling from one sector to an other. The shape of the
resulting sector is considered, as the simpler is
the shape, the easier it is to manage. Complex shapes are therefore avoided,
thanks to the notion of compactness in~\cite{jagare2013airspace} and balconies
in~\cite{sergeeva2017dynamic}. To smooth the transition between two
configurations, the work associated with the reallocation of one or more modules
is evaluated. This transition cost is included in the cost function being
minimised (in~\cite{bedouet2016towards}).

@@

The Monte Carlo tree search algorithm has been widely used in two-player games.
Only three years after its apparition in 1990, Br\"ugmann applies it to the Go
game in~\cite{brugmann1993mcgo}. While the Monte-Carlo tree search is still
extensively used in two-player games (and especially the Go game), its
adaptation to one player game has been worked on. Auer \textit{et al.} propose
an
upper confidence bound formula in~\cite{sebag2010fuse} which appears to be more
efficient for one player games. The introduced formula uses the standard
deviation of the outcome of the simulations.
The latter article also uses the rapid
action value estimation (RAVE) technique to quicken the convergence of the
algorithm. RAVE uses the ``all moves as first'' heuristic, in which all
moves\footnote{a move is informally considered as a decision taken regarding
which state to choose while descending the search tree}
seen during simulations are considered as a first move. This allows the
algorithm to update more statistics in one simulation. Gelly \textit{et al.} use
in~\cite{gelly2012grand} the RAVE technique coupled with several heuristics. The
heuristics bias the initialisaton of a node in the search tree, pre filling its
statistics using prior knowledge of the problem.

@@

The task of building an optimal sequence of partitions (lowering
as much as possible the workload of each controller) is fulfilled by a
stochastic tree search method, namely the Monte Carlo tree search. To assess the
quality of the results, an exact method (here A\(^*\)) is used.

@@

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then the MCTS algorithm is called back with
node \(v\)
as the root node. The final path is composed of the successive roots. The
algorithm is described in~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{\mu_v | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact best first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of terminal
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{N}^2 \to \mathbb{R}\). The
function \(\first(G)\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \first(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

@@

In this section, we discuss the model established to approach this configuration
problem. First we define a structure for a controlled sector and all possible
transitions through time. Then a workload model is defined. And finally we
define a cost per partition and a cost function aimed to be minimized.

@@

In the configuration problem, the airspace needs to be divided into sectors. Each
sector represents an area controlled by two controllers and is composed of
one or more elementary modules. A partition $P$ is a set of
non-overlapping sectors covering all the airspace.

A sector is a group of elementary modules, but in practice only a collection
of sectors are allowed in this model. This collection is referred to as the
context. It is the list of operation ATC sectors actually used in ATC centres.

@@

During the day, and with the evolution of the traffic, the workload (rigorously
defined in~\ref{sub:workload}) varies. To help controllers to maintain a
homogeneous level of workload, the airspace partitioning needs to be adapted
through successive transitions starting from an inital partition.
This maneuver requires either coordination or the opening of a new position
and hence increases controller's workload.
This is the reason why a full reconfiguration isn't feasible, and
only a subset of all available transitions can be operated.

In the model considered here, three transitions are possible namely a merge,
a transfer
or a split. For an airspace composed of three modules $A$, $B$, and $C$,
those actions represent:
\begin{itemize}
  \item merge: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A, B, C\}\}$
  \item split: $\{\{A, B, C\}\} \quad \rightarrow \quad \{\{A, B\}, \{C\}\}$
  \item transfer: $\{\{A, B\}, \{C\}\} \quad \rightarrow \quad \{\{A\}, \{B, C\}\}$
\end{itemize}

Each transition is composed of at most one merge, split or transfer; more would
result in too complex reconfigurations.

@@

Let us now precise the model established to evaluate a workload metric. The
workload depends on external criteria such as sector and traffic complexity
or human factors (e.g.\ controller health or stress). In the model assumed here,
the workload felt by a controller is directly and only related to the traffic
complexity.

The simplest approch considers only the number of aircraft per sector $N_\text{aircraft}$. We
define two arbitary thresholds, namely $th_\text{low}$ and $th_\text{high}$, defining
three zones where the workload is low (resp. normal and high) if $N_\text{aircraft} < th_\text{low}$
(resp. $th_\text{low} \leq N_\text{aircraft} \leq th_\text{high}$ and $th_\text{high} < N_\text{aircraft}$).
Those workload levels are translated to probabilities $p_\text{low}$, $p_\text{normal}$ and
$p_\text{high}$, where $p_\text{low}(S)$ is the probability that the sector $S$ is underloaded
(resp. balanced and overloaded). In short, for a sector $S$ at time $t$ and $N_\text{aircraft}$:
\begin{equation}
  \begin{aligned}
    p_\text{low}^{S, t} &= \mathbb{1}_{[0, th_\text{low}]}(N_\text{aircraft}) \\
    p_\text{normal}^{S, t} &= \mathbb{1}_{[th_\text{low}, th_\text{high}]}(N_\text{aircraft})\\
    p_\text{high}^{S, t} &= \mathbb{1}_{[th_\text{low}, \infty[}(N_\text{aircraft})
  \end{aligned}
\end{equation}
where $\mathbb{1}$ is the indicator function defined by:
\begin{equation}
  \mathbb{1}_{I}(x) =
  \begin{cases}
    1 & \text{if\,} \quad x \in I\\
    0 & \text{otherwise}
  \end{cases}
\end{equation}

This expression of the workload, expressed in terms of probabilities, allows us to
use a more complex model for the workload such as a neural network
(see~\cite{gianazza2010forecasting}). This method learns the probabilities $p_\text{low}$,
$p_\text{normal}$ and $p_\text{high}$ based on parameters including traffic complexity and
the complexity of the sector (e.g. airspace volume).

@@

In order to evaluate a given partition $P$ at time $t$, a cost $C(P, t)$ needs
to be defined. In this paper, this cost depends on the workload in each
sector. The definitions given in~\cite{ferrari2017} are used to represent a high
(respectively normal and low) cost for partition \(P\):
\begin{equation}
  \begin{aligned}
    c_+(P, t) &= \sum_{S \in P_t} \delta_{h}(S, t) \cdot p_\text{high}^{S, t} \cdot |S|^2\\
    c_=(P, t) &= \left(
    \sum_{S \in P_t} \delta_{n}(S, t) \cdot p_{norm}^{S} \cdot |S|^{-2}
    \right)^{-1}\\
    c_-(P, t) &= \sum_{S \in P_t} \delta_{l}(S, t) \cdot p_\text{low}^{S, t} \cdot |S|^{-2}
  \end{aligned}
\end{equation}
with $\delta_{h}(P, t)$ (resp. $\delta_{n}(P_t)$ and $\delta_{l}(P_t)$) equals 1
if the probability $p_\text{high}$ (resp. $p_\text{normal}$ and $p_\text{low}$) is superior to
the two others, and 0 otherwise.

@@

The parameters $\alpha$, $\beta$, $\gamma$ and $\lambda$ (all positive)
determine a priority on which parameter to optimize. For instance, a high
value $\alpha$ help to
minimize $c_+$, hence the overall number of sectors with too much traffic.
In a real application, it maybe interesting to order those parameters as
follow: $\alpha > \gamma > \beta > \lambda$.

@@

with $\theta > 0$ a parameter to determine.
This is the objective function to minimize in the A$^*$ algorithm.

@@

The traffic is simulated the following way, each elementary module has an
associted list which, at index \(i\) has the number of aircraft in it at time
step \(i\). Those lists have been randomly generated. The set of lists form a
scenario which is used in all following tests.

@@

Considering paragraph~\ref{ssec:theoretical_performances}, the MCTS should
eventually converge to the optimal value. This result reveals a flaw in the
implementation.
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/longterm.data};
      \addplot+[dashed, domain=0:10] {1268};
    \end{axis}
  \end{tikzpicture}
  \caption{The continuous line shows the total cost of a path computed by the
    MCTS algorithm. The x-axis shows the time allowed to the algorithm. The
    dashed curve is 15\% of the optimal cost found with the \(A^*\)
  algorithm.}\label{fig:longterm}
\end{figure}
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}
      [line width=0.05, mark size=0.1, xlabel=time(s), ylabel=cost]
      \addplot+ table {data/shortterm.data};
      \addplot+[dashed, domain=0:2] {1268};
    \end{axis}
  \end{tikzpicture}
  \caption{}\label{fig:shortterm}
\end{figure}
